---
title: "R4MSWorkshop2"
format: html
editor: source
---

## Starting workshop 2

Loading things in

```{r}
#install.packages('tidyverse')
library('tidyverse')
table1
table2
table3
```

2.2 tidying data

```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)
# Compute cases per year
table1 %>% 
  count(year, wt = cases)
# Visualise changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

2.4 Pivoting data to make it tidy+lengthening datasets With Billboard data

```{r}
billboard
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )

df <- tribble(
  ~id,  ~bp1, ~bp2,
  "A",  100,  120,
  "B",  140,  115,
  "C",  120,  125
)

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )

```

2.5 Widening datasets #using medical dataset

```{r}
cms_patient_experience
cms_patient_experience |> 
  distinct(measure_cd, measure_title)

cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )

cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
print(cms_patient_experience)
```

Pivoting wider #bloodPressure example

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

df |> 
  distinct(measurement) |> 
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

#pivot wider 
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)

  
```

2.6 Exercise with palmerpenguins

```{r}
library(palmerpenguins)
# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)

#visualize the dataset
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

#calc mean body mass for each species on each island
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

#above is good but we can use pivot_wider to tidy data even better for publication
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )

head(mass_matrix)


```

2.7 separting and uniting datasets #using table 3 dataset

```{r}
table3
#split column up into 2 variables (cases and pop)
table3 %>% 
  separate(rate, into = c("cases", "population"))

#if you want to separate using a specific character
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")


#now variables listed as interger types with true value
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

#separates the last two digits of each year
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

#use unite to combine multi columns 
table5 %>% 
  unite(new, century, year, sep = "")
```

2.8 Wrangling strings and dates

```{r}
library(tidyverse)
#help sorting messy datesets with different types of the same data/name

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(

# Convert everything to lowercase
site_clean = str_to_lower(site_id),
# Trim any leading or trailing whitespace (invisible spaces at the ends)
site_clean = str_trim(site_clean),
# Replace any spaces with underscores
site_clean = as_factor(str_replace_all(site_clean, pattern = " ", replacement = "_")))

# 

  
str(clean_sites)
 
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2
 
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)


    
```
2.9 relational data: Joining tables #fish data 
```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)
#innerjoin only keeps matches that are completely filled
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)

#using anti_join to find unmatched rows
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))



```

2.10 Handling missing values #temperature data
```{r}
# Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

#replacing missing values (0 that turn in to NA) with coalese
#shark data
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)


#NAN Not a Number Scenario
#catch per unit data\
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)

#missing values and 0 framwork
#coral Trout Data
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))

# Force inclusion of hidden zero-catch data
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)

#getting rid of NA, ex. na =failed reading on thermometer
#salinity 
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)

print(clean_log)

```
Load helping tools 
```{r}
library(chattr)
library(ellmer)

# Create the Gemini chat connection
chat <- ellmer::chat_google_gemini(model = "gemini-flash-latest")

# Tell chattr to use this connection
chattr_use(chat)
```

2.11 Practical exercises: messy data sets Penguins 
```{r}
# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)
library(lubridate)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)

#exercise 1 cleaning up the metadata
clean_metadata  <- island_metadata |>
  mutate(
    # Clean up island names
    island_name = str_trim(island_name),
    island_name = str_to_title(island_name),
    #lubricate the dates to a proper format
    date_format = case_when(
      str_detect(station_install, "/") ~ "dmy",
      str_detect(station_install, "-") ~ "dmy",
      TRUE ~ "unknown"
    ),
    # Convert station_install to a proper date
    station_install = dmy(station_install)
  )
print(clean_metadata)

# exercise 2 combine clean metadata with penguins dataset
penguins_spatial  <- palmerpenguins::penguins |>
  left_join(clean_metadata, by = join_by(island ==island_name))

#scroll to the right to see the new columns 
head(penguins_spatial)

#exercise 3 clean table showing maximum body mass recorded for each species, separted by island 
clean_summary <- penguins_spatial |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(max_mass = max(body_mass_g), .groups = "drop") |>
  pivot_wider(
    names_from = island,
    values_from = max_mass
  )
print(clean_summary)
```

