---
title: "WS2_Estuary_Prac"
# Output Config
format: 
  # HTML Output
  html:
    theme: cosmo          # Modern Bootstrap theme with clean typography
    highlight-style: a11y # High-contrast syntax highlighting for code
    toc: true             # Includes a Table of Contents
    toc-depth: 3          # Captures down to sub-subsections (e.g., 3.1.1)
    toc-location: left    # Pins TOC to left sidebar so it doesn't block main page
    toc-title: "Contents" # Custom title for the TOC sidebar
    number-sections: true # Numbers sections automatically (e.g., 1, 1.1, 2)
    code-fold: true       # Collapses code blocks by default under a dropdown
    code-summary: "Show code" # Label text on collapsed code blocks
    code-tools: true      # Adds global code menu (Show/Hide All Code) at top right
    code-copy: true       # Adds a one-click copy button to code blocks
    embed-resources: true # Bundles images/CSS into a single stand-alone HTML file
    df-print: paged       # Formats data frames into interactive paged tables
    fig-height: 5         # Default plot height in inches
    fig-width: 8          # Default plot width in inches
    #fig-responsive: true # Scales figures smoothly across screen sizes
    smooth-scroll: true   # Enables smooth scrolling when clicking TOC links
    
# Code Execution Defaults across all chunks
execute:
  echo: true              # Display source code alongside outputs
  eval: true              # Execute the code in the chunk
  include: true           # Include the chunk in the final rendered document
  warning: false          # Suppress warning messages in final rendered document
  message: false          # Suppress package/function load messages in rendered output
  knitr:
    opts_chunk: 
      comment: "#>"       # Clean output prefix for console printed output
      dev: "cairo_pdf"    # Uses high-quality Cairo device for crisp vector graphics in PDF

---
Estuary Practical


# Housekeeping
```{r}
#| message: true
#| warning: false
rm(list = ls())
```

#loading in R packages 
```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

#load excel files
```{r}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

#inspect indiviual tabs
```{r}
catch1
catch2
catch3
catch4
```

#bind rows together
```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

#prelim check
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#check specific variables
```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```
# Load metadata
```{r load-metadata}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
```
# Checks
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```

### Fix site names in data and metadata and convert character to factor variables
```{r fix-site-names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```
# Recheck
```{r recheck-site-names}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
str(metadata)
```
###  Load Taxonomic Dictionary
```{r load-species-dict}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
# Check
head(species_dict)
```
#check species name
```{r}
catch_data |> 
  distinct(species)
# Can see these are different to the dictionary
species_dict |>
  distinct(common_name)
```
#fix species names in catch_data to match dictionary
```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```
### Create new variable ("scientific_name") in catch log by joining with taxonomic dictionary 
```{r join-species-dict}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```
#recheck
```{r}
glimpse(catch_data)
```
# Load the sonde data, given code for data
```{r load-sonde}
source(here::here("code/diw_sensor.R"))
```
#join catch data with metadata and sensor data
```{r}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```
#check
```{r}
glimpse(catch_data_final)
```
#wrangle data set to include the zeros
```{r}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```
###Finished data wrangling####

#Look at values
#### Salinity
```{r}
# Let's make a table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
   # sem_salinity = sd(mean_salinity) / sqrt(n()),
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

# And from this a plot
sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
```
### Turbidity
```{r}
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Turbidity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Turbidity (NTU)")
```
### And temperature
```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Temperature by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Temperature (°C)")

```
#### Remake salinity plot nicely for publication
```{r}
#| fig-cap: "Salinity Gradient across zones. Ross River Estuary System, Townsville, NQ, Australia May 2026."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(#title = "Median Salinity by Zone with Standard Deviation Error Bars",
       x = "Zone",
       y = "Median Salinity (ppt)")
  
```
# Calculate summary stats #fish abundance
```{r}
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )
```

#plot of fish abundance
```{r}
fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = scientific_name)) +
  geom_line(group = 1) +
  # add in points from master data
  geom_point(data = catch_data_final, aes(x = zone, y = count),
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) + 
  geom_point(position = "dodge", size = 4, pch = 23) +
  # SE pretty narrow, potentially not worth plotting but let's viewer see that included
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), 
                position = position_dodge(0.9), width = 0.2) + 
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"), # spp names in italic
   # legend.position = "bottom"
    legend.position = "none" # legend unnecessary
  ) +
  labs(
    #title = "Predatory Fish Abundance across Estuarine Salinity Gradients", # default ggplot title appears on top
    #subtitle = "Ross River Estuary System, May 2026",
    x = "Spatial Zone",
    # scientific label for y axis of n day -1 use regex to get superscript
    y = expression("Abundance (n day"^-1*")"),
  #  y = "Daily abundance (n day^-1)"
    y = "Abundance (per day)"
  )
```
```{r}
ggsave("fish_abundance_plot.png", width = 10, height = 6, units = "in", dpi = 300)
```

