---
title: "WS2_Estuary_Prac"
format: html
editor: visual
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---
Estuary Practical
```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Housekeeping
```{r}
#| message: true
#| warning: false
rm(list = ls())
```

#loading in R packages 
```{r}
library(tidyverse) # Inc dplyr, ggplot2, lubridate etc
library(readxl) # For reading in excel files
```

#load excel files
```{r}
catch1 <- read_excel(here::here("data/workshop2/estuary_catch_log.xlsx"), sheet = "Lower_ross")
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

#inspect indiviual tabs
```{r}
catch1
catch2
catch3
catch4
```

#bind rows together
```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

#prelim check
```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#check specific variables
```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```
# Load metadata
```{r load-metadata}
metadata <- read_csv("../data/workshop2/estuary_metadata.csv")
```
# Checks
```{r inspect-metadata}
head(metadata)
glimpse(metadata)
```

### Fix site names in data and metadata and convert character to factor variables
```{r fix-site-names}
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)),
    site = str_replace_all(site, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )

metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  # Convert all chr to fct
  mutate(across(where(is.character), as_factor)
  )
```
# Recheck
```{r recheck-site-names}
catch_data |> 
  str()
# catch_data |> 
#   distinct(site)
str(metadata)
```
###  Load Taxonomic Dictionary
```{r load-species-dict}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")
# Check
head(species_dict)
```
#check species name
```{r}
catch_data |> 
  distinct(species)
# Can see these are different to the dictionary
species_dict |>
  distinct(common_name)
```
#fix species names in catch_data to match dictionary
```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )
```
### Create new variable ("scientific_name") in catch log by joining with taxonomic dictionary 
```{r join-species-dict}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))
```
#recheck
```{r}
glimpse(catch_data)
```
# Load the sonde data, given code for data
```{r load-sonde}
source(here::here("code/diw_sensor.R"))
```
#join catch data with metadata and sensor data
```{r}
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |>
  left_join(sensor_daily, by = join_by(site, date))
```
#check
```{r}
glimpse(catch_data_final)
```
#wrangle data set to include the zeros
```{r}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```
###Finished data wrangling####



